---
title: "Tidy Data day 2"
output: github_document
---

# Homework Solution:

```{r, echo=F}
tycho_timeOnly_link <- "https://raw.githubusercontent.com/UNC-HLC-R-Intro/lessons/import/data/tycho_wide_timeOnly.csv?token=AQdxTdVbMVfOF85YawF5KGZrS_i0xTxDks5bUKzswA%3D%3D"

tycho <- readr::read_csv(tycho_timeOnly_link)
```

Specify names of melted columns with reshape
```{r}
tycho_melt <- tycho %>% 
  reshape2::melt(id.vars = c("state", "disease"), value.name = "incidence", variable.name = "yearweek") 
```

Convert yearweek from **factor** to **numeric**.

With `as.character(yearweek) %>% as.numeric`

You have to do it this way because R sees factors as numbers first unless you say otherwise. 

`yearweek %>% as.numeric` would output `1,2,3,4,5...` because these are the factor **levels**

The measles vaccine came out in 1960, so that yearweek code is 196001
```{r}
tycho_melt %>% 
  dplyr::filter(disease == "MEASLES") %>% 
  dplyr::mutate(yearweek = as.character(yearweek) %>% as.numeric) %>% 
  ggplot(aes(yearweek, incidence)) +
    geom_line(aes(group = disease)) +
    ggtitle("Measles over time in USA") +
    geom_vline(xintercept = 196001, color = "red")
```


# Day2:

Today we will work with the same data, but in a slightly more difficult format.



```{r}
tycho_wide_timeDisease_link <- "https://raw.githubusercontent.com/UNC-HLC-R-Intro/lessons/import/data/tycho_wide_time-disease.csv?token=AQdxTfHxKKtnxxzNvQeWm5VXTBq9gZ0-ks5bUKzQwA%3D%3D"
```

```{r}
tycho <- readr::read_csv(tycho_wide_timeDisease_link)
```

# `tidyr` package for more complicated data manipulation

```{r}
tycho %>% 
  head
```

The goal of today is to reproduce the above plot from this more complicated data
structure. We'll be using another package from the `tidyverse` called `tidyr`
([documentation](https://tidyr.tidyverse.org/reference/index.html))

One good way to think about reshaping data is to work *backwards* from what you want, and come up with the steps to get to the end.

We want our data to look like this:
```{r, echo=F}
library(reshape2)
tycho %>% 
 melt() %>% head %>% 
 tidyr::separate(variable, c("yearweek", "disease"))
```

So we should make data that first looks like this:
```{r, echo=F}
tycho %>% 
  melt() %>% head
```

And then separate the `variable` column into two new columns, `yearweek` and `disease`.

It is very easy to separate two values from eachother if they are delimited by a separator character (in this case "_").

The `tidyr` package provides several functions for manupulating data frames that are in tidy format.

For more reading: 
- [R for Data Science Chapter](http://r4ds.had.co.nz/tidy-data.html)

- [Tidyr tutorial](https://www.r-bloggers.com/data-manipulation-with-tidyr/amp/)


## Exercise 1
Melt `tycho` to make it tidy.

```{r, echo = F}
tycho %>% 
  melt(value.name = "incidence")
```

## Exercise 2

Read the documentation for the `tidyr` function `separate`. 
Create the new columns `yearweek` and `disease` from the melted data above.

```{r, echo = F}
tycho %>% 
  melt(value.name = "incidence") %>% 
  separate(variable, c("yearweek", "disease")) %>% 
  head
```


