---
title: "Functions, for loops, and if/else constructs"
author: "Jamshaid Shahir"
date: "June 17, 2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Basic functions
So far we've been using a lot of functions that exist within R. Now we'll learn how to build our own function.


```{r eval=FALSE}
myfunction <- function(arg1, arg2, ...) {
  
  #do some stuff
  return(object)
  
}
```

Make a simple function that just returns the argument that it is given.

```{r}
first_function <- function (x) {
  
  return(x)
  
}
```

Now if you pass a value to `first_function` it should return that value:

```{r}
first_function(9)
```

A function that only returns the given value is not useful, but it does illustrate how you can add functions to your environment. Now make a function that squares a value and then adds one to it.

```{r}
second_function <- function(x) {
  ans <- x^2 + 1
  return(ans)
  
}
second_function(9)
```


### Exercise 1

Suppose we wanted to look at the Gapminder dataset and calculate GDP. Write a function to do that, returning the dataset with the new GDP column
```{r, echo=F, message=F, warning=F}
library(dplyr)
library(gapminder)
```

```{r, echo=F, eval=F, include=F}
library(dplyr)
library(gapminder)

getGDP <- function(data)
{
  new_data <- mutate(data, GDP=gdpPercap*pop)
  return(new_data)
}

getGDP(gapminder)

# Alternate answer:
# Makes general function that computes GDP from 2 columns, which can be passed to mutate

getGDP2 <- function(data){
  gdp <- data$gdpPercap * data$pop
  return(gdp)
}

# Returns a vector of GDPs
getGDP2(gapminder)

# Inside mutate will capture the vector in a column named `GDP`
gapminder %>% 
  dplyr::mutate(GDP = getGDP2(.))

```

### Exercise 2

Write a function that converts a Farhenheit temperature to Kelvin (remember that K = 5/9*(F - 32) + 273.15).

```{r, echo=F, eval=F, include=F}
f2k <- function(fah) {
  kelvin <- (5/9)*(fah - 32) + 273.15
  return(kelvin)
}

f2k(60)
```

### For loops

For loops are used to iterate through elements in a vector, repeating the same block of code each time. They take the basic form:

```{r, eval=FALSE}
for(i in vector) {
  Do some operation to each element of the vector
}
```

Example:

```{r}
for (year in c(2010,2011,2012,2013,2014,2015)){
  print(paste("The year is", year))
}
```

We can also store the output of the code block evaluated in each iteration of the for loop:

```{r}
output <- c() # create an empty vector
for (year in c(2010,2011,2012,2013,2014,2015)){
  output <- c(output, paste("The year is", year)) # add to the empty vector each iteration
}
output
```

We can also do this another way by iterating through the indices with the `length` function. This is rather common in more traditional programming languages, such as MATLAB, C++, and Java.

```{r}
output <- c() # create an empty vector
years <- c(2010,2011,2012,2013,2014,2015)
for (i in 1:length(years)){
  output[i] <- years[i] # add to the empty vector each iteration
}
output
```
### Exercise 3

Use a for loop to print out all the countries in the Americas from Gapminder (hint: use `filter` to just extract the American countries; also, how can you avoid repeats?)

This usually is not preferred if you want to keep the output of a for loop, because it can become inefficient. Instead, we could use a function from the `apply` family.

```{r, echo=F, eval=F, include=F}
us_Gapminder <- filter(gapminder, continent == "Americas" & year == 2002)

for (i in 1:length(us_Gapminder$country))
{
  print(as.character(us_Gapminder$country[i]))
}
```

The levels tacked on make our output a little hard to read. We can suppress those by using the following function: `as.character`

```{r}
head(as.character(gapminder$continent))
```


## For loop exercise

Use the following vector `years` and the function `printYear` inside a for loop to print a new line for each year in `years`.

```{r}
years <- c(2010,2011,2012,2013,2014,2015)

printYear <- function(year) {
  paste("The year is", year)
}

```


### If/else statements

In R, you create an if/else statement with the following syntax:

```{r,eval=FALSE}
if (logical statment) { do something } else  { do something different } 
```

For example:

```{r}
if (1 > 2) { print("TRUE") } else { print("FALSE") }
```

Not all if/else statements have to have an `else` clause. By default, if no `else` statement is given R does nothing.

```{r}
x <- 1000
if (x > 100) print("'x' is greater than 100")
```

If/else statements can also be nested:

```{r}
y <- 25
if (y < 100) {
  if (y > 10) {
    print("10 < y < 100")
  } else {
    print("y < 10 or y > 100")
  }
} else {
  print("y < 10 or y > 100")
}
```

### Exercise 5

Without using `filter`, loop through the entire Gapminder dataset, and just print out countries in the American continent. For simplicity and to avoid repeats, just use data from the year 2002: `gap2002 = filter(gapminder, year==2002)`
```{r, echo=F, eval=F, include=F}
gap2002 <- filter(gapminder, year==2002)

for (i in 1:length(gap2002$continent))
{
  if (gap2002$continent[i] == "Americas") 
  {
    print(as.character(gap2002$country[i]))
  }
}
```

### Bonus

Write a function that takes as input a gapminder dataset from 2002 and a specified continent, and prints out the list of countrys

```{r, echo=F, eval=F, include=F}
printCty <- function(data, cont)
{

  for (i in 1:length(data$continent))
  {
    if (data$continent[i] == cont) 
    {
      print(as.character(data$country[i]))
    }
  }
}

printCty(gap2002, "Americas")
printCty(gap2002, "Africa")
printCty(gap2002, "Asia")
```


### Apply family functions

Apply family functions similarly perform operations iteratively on slices of data (rows in a matrix, elements in a list). For many operations, the apply family can be more efficient than a for loop. You can read on about apply functions at [Data Camp](https://www.datacamp.com/community/tutorials/r-tutorial-apply-family#family).

We can use `sapply`, which applies a function over a list or vector and returns the output as a vector, to get output as the above for loop:

```{r}
years <- c(2010,2011,2012,2013,2014,2015)
sapply(years, function(x) paste("The year is", x))
```

We can use `lapply`, which applies a function over a list or vector and returns the output as a list.

```{r}
years <- c(2010,2011,2012,2013,2014,2015)
lapply(years, function(x) paste("The year is", x))
```

### Exercise 4

Repeat the prior exercise using `sapply` and `lapply`
```{r, echo=F, eval=F, include=F}
us_Gapminder <- filter(gapminder, continent == "Americas" & year == 2002)

sapply(us_Gapminder$country, function(x) x)
lapply(us_Gapminder$country, function(x) x)
```