---
title: "Functions day 2"
output: github_document
---

### For loops

For loops are used to iterate through elements in a vector, repeating the same block of code each time. They take the basic form:

```{r, eval=FALSE}
for(i in vector) {
  Do some operation to each element of the vector
}
```

Example:

```{r}
for (year in c(2010,2011,2012,2013,2014,2015)){
  print(paste("The year is", year))
}
```

We can also store the output of the code block evaluated in each iteration of the for loop:

```{r}
output <- c() # create an empty vector

for (year in c(2010,2011,2012,2013,2014,2015)){
  output <- c(output, paste("The year is", year)) # add to the empty vector each iteration
}

output
```

This usually is not preferred if you want to keep the output of a for loop, because it can become inefficient. Instead, we could use a function from the `apply` family.

### Apply family functions

Apply family functions similarly perform operations iteratively on slices of data (rows in a matrix, elements in a list). For many operations, the apply family can be more efficient than a for loop. You can read on about apply functions at [Data Camp](https://www.datacamp.com/community/tutorials/r-tutorial-apply-family#family).

We can use `sapply`, which applies a function over a list or vector and returns the output as a vector, to get output as the above for loop:

```{r}
years <- c(2010,2011,2012,2013,2014,2015)

sapply(years, function(x) paste("The year is", x))
```

#### Exercise 1

Use `sapply` to replace all of the 20s in `years` with 30s (think back to `gsub` from the regex section of the data import lesson).

### If/else statements

In R, you create an if/else statement with the following syntax:

```{r,eval=FALSE}
if (logical statment) { do something } else  { do something different } 
```

For example:

```{r}
if (1 > 2) { print("TRUE") } else { print("FALSE") }
```

Not all if/else statements have to have an `else` clause. By default, if no `else` statement is given R does nothing.

```{r}
x <- 1000
if (x > 100) print("'x' is greater than 100")
```

If/else statements can also be nested:

```{r}
y <- 25
if (y < 100) {
  if (y > 10) {
    print("10 < y < 100")
  } else {
    print("y < 10 or y > 100")
  }
} else {
  print("y < 10 or y > 100")
}
```

#### Exercise 2

Use `genPrimer` from functions day 1 to generate a list of 20 primers that are between 20-30 base pairs in length.

Write a for loop that prints "primer is between 20 and 24 bases" if the primer has 20 or more bases and 24 or fewer bases. 

#### Exercise 3

Make a vector of the days of the week:

```{r}
days <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
```

Use sapply to create a vector that contains abbreviations for the days of the week using days. Use an if/else statement to keep just the first character of each day of the week that starts with a unique letter (Monday, Wednesday, Friday), but keep two characters for days that start with the same letter as another day (Tuesday, Thursday, Saturday, and Sunday).

Hint: use the function substr to extract a subset of the characters from a character string.
