---
title: "Functions day 1"
output: github_document
---

### Basic functions
So far we've used a lot of functions that already exist in R. Now you will learn how to write your own functions. User-written functions take the following structure:

```{r eval=FALSE}
myfunction <- function(arg1, arg2, ...) {
  
  do some stuff
  return(object)
  
}
```

Make a simple function that just returns the argument that it is given.

```{r}
first_function <- function (x) {
  
  return(x)
  
}
```

Now if you pass a value to `first_function` it should return that value:

```{r}
first_function(9)
```

A function that only returns the given value is not useful, but it does illustrate how you can add functions to your environment. Now make a function that squares a value and then adds one to it.

```{r}
second_function <- function(x) {

  ans <- x^2 + 1
  return(ans)
  
}
second_function(9)
```

#### Exercise 1

Write a function that converts a Farhenheit temperature to Kelvin (remember that K = 5/9*(F - 32) + 273.15).

#### Exercise 2

Write a function that returns the length and sum of a given vector of numbers. How can you work around the fact that `return` can only be used once in a function?

#### Exercise 3

The following function generates random primers. Use it to create a list of 50 primers between 40-50 base pairs:

```{r}
library(stringi) # needed for function stri_rand_strings

genPrimer <- function(minlen, maxlen, nprimers, bases = '[ACTG]') {

  lengths <- lengths1 <- sample(minlen:maxlen, nprimers,replace=TRUE)
  res <- stri_rand_strings(nprimers, lengths, bases)
  return(res)
  
}
```

Then, write a function that counts the number of G's in each of your randomly generated primers from above, and returns three items: the primer sequence, the primer length, and the number of G's in that primer.  

*Hint:* Use the function `str_count` in the library `stringr` to count the G's.
