---
title: "Functions Part 2: Control Statements"
author: "Amy Pomeroy, Dayne Filer, & Sara Taylor"
date: "May 30, 2018"
output: 
  pdf_document:
    highlight: tango
  html_document: null
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
set.seed(1234)
```

<!-- Here we style out button a little bit -->
<style>
  .showopt {
    background-color: #004c93;
    color: #FFFFFF; 
    width: 100px;
    height: 20px;
    text-align: center;
    vertical-align: middle !important;
    border-radius: 8px;
    float:right;
    }

    .showopt:hover {
    background-color: #dfe4f2;
    color: #004c93;
  }

</style>

<!--Include script for hiding output chunks-->
<script src="hideOutput.js"></script>


### Using this document
* Code blocks and R code have a grey background (note, code nested in the text is not highlighted in the pdf version of this document but is a different font).
* \# indicates a comment, and anything after a comment will not be evaluated in R
* The comments beginning with \#\# under the code in the grey code boxes are the output from the code directly above; any comments added by us will start with a single \#
* While you can copy and paste code into R, you will learn faster if you type out the commands yourself. 
* Read through the document after class. This is meant to be a reference, and ideally, you should be able to understand every line of code. If there is something you do not understand please email us with questions or ask in the following class (you're probably not the only one with the same question!). 

### Expectations
1. Implement the three basic control statements in R (for-loops, if/else statements, and while statements)
2. Use the and/or operators for combining logical statements

### Control statements
Imagine you have a data set of primer sequences. You want to quickly find out information about this set of primers. Let's start by loading the list.

```{r}
setwd("C:/Users/gdicecco/Desktop/git/rclass/")
primers <- read.csv("primers.csv", header = F, stringsAsFactors = F)
str(primers)
```

There are 562 primers in the list, so if you wanted to find the length of all of them or only primers longer than a certain length, you probably wouldn't want to count the length of each one by hand. Control statements can help us make these tasks quicker. 

### If/else statements
In R, you create an if/else statement with the following syntax:

```{r,eval=FALSE}
if (logical statment) { do something } else  { do something different } 
```

For example:

```{r}
if (TRUE)  { print("TRUE") } else { print("FALSE") }
if (FALSE) { print("TRUE") } else { print("FALSE") }
if (1 > 2) { print("TRUE") } else { print("FALSE") }
```

Not all if/else statements have to have an `else` clause. By default, if no `else` statement is given R does nothing.

```{r}
x <- 1000
if (x > 100) print("'x' is greater than 100")
```

If/else statements can also be nested:

```{r}
y <- 25
if (y < 100) {
  if (y > 10) {
    print("10 < y < 100")
  } else {
    print("y < 10 or y > 100")
  }
} else {
  print("y < 10 or y > 100")
}
```

__$\color{blue}{\text{Exercise 1}}$__
Write a for loop that prints "primer is longer than 24 bases" if the primer is longer than 20 bases, "primer is exactly 24 bases" if the primer is 20 bases, and "primer is shorter than 24 bases" if the primer is shorter than 24 bases.

### Combining if/else statments

Another way to combine if/else statements is to combine the conditional statement. There are four operators for combining logical statements: `&`, `&&`, `|`, `||`. The "and" operators (`&`/`&&`) return return `TRUE` if both statements are `TRUE`. The "or" operators (`|`/`||`) return `TRUE` if either statement is `TRUE`. The single-character operators (`&` and `|`) return a vector. 

```{r}
1 > 0 & TRUE
FALSE | 10 ## Remember from class 1 how 10 is coerced to TRUE
1 > 0 & c(TRUE, FALSE, FALSE, TRUE)
0 > 1 & c(TRUE, FALSE, FALSE, TRUE)
0 > 1 | c(TRUE, FALSE, FALSE, TRUE)
```

The double-character operators (`&&` and `||`) will only return a single `TRUE` or `FALSE` value. 

```{r}
1 > 0 && 10 < 100
1 > 0 && FALSE
1 > 0 || FALSE
1 > 0 && c(TRUE, FALSE)
1 > 0 && c(FALSE, TRUE)
```

Notice in the last two statements above the order of the vector matters. The double-character operators are lazy -- they only check as much as they need to. If you give a double-character operator a vector, it only checks the first element. 

Furthermore, if the first expression is `TRUE` the `||` operator will not even evaluate the second expression. Similarly, if the first expression is `FALSE` the `&&` operator will not evaluate the second expression.

```{r,error=TRUE}
TRUE  || r
TRUE  && r
FALSE || r
FALSE && r
```

You can combine logical expressions within an if/else statement (recall we defined `y` as `25` in an earlier expression). 

```{r}
if (y < 100 && y > 10) {
  print("10 < y < 100")
} else {
  print("y < 10 or y > 100")
}
```

__$\color{blue}{\text{Exercise 2}}$__
Write a for loop that prints "primer is between 20 and 24 bases" if the primer has 20 or more bases and 24 or fewer bases. 

__$\color{blue}{\text{Exercise 3}}$__
Write a script that calculates the melting temperature (Tm) of each primer. For primers that are 13 bases or less the formula for Tm in Celcius is: $Tm= (A+T) * 2 + (G+C) * 4$, where A, T, G, and C represent the number of each base.  For primers that are 14 bases or more the formula for Tm in Celcius is: $Tm= 64.9 +41*(G+C-16.4)/(A+T+G+C)$. 

### While statements
The final control statement we will discuss are while statements. Although while statements are rarely used in R, they are often used in other languages and are good to understand. The while syntax in R is:

```{r,eval=FALSE}
while (condition) {
  do something
  update condition
}
```

Try the simple example below.

```{r}
val <- 3
while (val < 10) {
  print(val)
  val <- val + 1
}
```

If you omit the second part that updates `val` R would have printed `3` until you manually killed the process.

### The apply function
The `apply` function is part of the `apply` family of functions in R, which are used to repetitively manipulate slices of data from structures such as matrices, lists, and dataframes and often can be used in place of loops. Here, we look at the `apply` function, which operates on arrays (a matrix is a 2D array). The arguments of `apply` are 1) an array, 2) the dimension over which the function is applied (`MARGIN = 1` indicates over rows, `MARGIN = 2` indicates over columns), and the function to apply (summarizing, subsetting, or user-defined).

In the following example, the `apply` function is used to take sums of each column in a matrix:

```{r}
m <- matrix(rnorm(20), nrow = 4, ncol = 5)
apply(m, MARGIN = 2, sum)
```

__$\color{blue}{\text{Exercise 4}}$__
Use `apply` to determine the length of each primer in the dataset `primers`. Try to find the function for counting the number of characters in a string on your own using the Help tab in RStudio or Google.

### Combining control statements
The following example combines for loops and if/else statements. Let's revisit our primer list example. We now want to calculate the molecular weight of each of our primers. We know that the molecular weights dCTP, dATP, dGTP, and dTTP are 467.2, 491.2, 507.2, and 482.2 g/mol respectively. 

```{r}
masses <- numeric(length=length(primers$V1)) #make empty numeric vector with the same length as the primers vector
for (i in 1:length(primers$V1)){
  plen <- nchar(primers[i,1]) #get length of ith primer
  pmass <- 0 #set mass of primer equal to zero
  for (j in 1:plen){
    l <- substr(primers[i,1],j,j) #get jth nucleic acid of ith primer
    if (l=="C"){
      m <- 467.2
    } else if (l=="A"){
      m <- 491.2
    } else if (l=="G"){
      m <- 507.2
    } else if (l=="T"){
      m <- 482.2
    } #get mass of jth nucleic acid of ith primer
    pmass <- pmass+m #add mass of jth nucleic acid to the mass of the primer
  }
  masses[i]<-pmass #store the mass of the ith primer in the vector "masses"
}
```

You can now look at the results: 
```{r}
masses
```