---
title: "Data Import & Cleaning"
date: "25 June, 2019"
output: github_document
---

```{r}
library(ggplot2)
library(dplyr)
library(readr) # Functions for importing & exporting files
```


R can read files from your computer, or it can use a weblink to a file. Today
we'll read in this file I've put up on github so it's easier to distribute to
all of you. If you wanted to read in your own data, you would give R the path to
your file on your machine.









```{r data.creation, echo=F}
people <- tibble::tribble(~"name", ~"age", ~"favorite_color",
                "Alice", 20, "Blue",
                "Bob", 18, "Red",
                "Claire", 23, "Blue",
                "Doug", 21, "Orange",
                "Elizabeth", 18, "Green")

readr::write_csv(people, "people.csv")
```

```{bash, echo = F}
echo "# This dataset contains information about people, their ages, and their favorite colors" > people2.csv
```
```{r, echo =F}
#readr::write_csv(people, "people2.csv", append = T, col_names = T)
readr::write_csv(people, "people2.csv", append = T)
```

```{r, echo=F}
readr::read_csv("people2.csv", comment = "#")
```

## People dataset
```{r, eval=F}
# These are 2 different data files hosted on Github. R can stream files from web links (not usually best way to get data).
people_link <- "https://raw.githubusercontent.com/UNC-HLC-R-Intro/lessons_2019/master/data_import/people.csv"
people2_link <- "https://raw.githubusercontent.com/UNC-HLC-R-Intro/lessons_2019/master/data_import/people2.csv"
```

```{r, echo=F}
# This let notebook knit incase github links expire
people_link <- "people.csv"
people2_link <- "people2.csv"
```

```{r}
people <- read_csv(people_link)

people
```

```{r}
read_csv(people2_link)
```

```{r, echo=F, eval=F}
# This dataset must ignore comments & you have to give it col_names
read_csv(people2_link, comment = "#")
```


```{r, echo=F, eval=F}
# complete version
read_csv(people2_link, comment = "#", col_names = c("name", "age", "favorite_color"))
```

### dplyr can rename columns
```{r}
people_renamed <- people %>% 
  rename(color = favorite_color)
```

### Writing files
`readr` has several `write_<type>` functions.

```{r}
write_csv(people_renamed, path = "people_renamed.csv", col_names = T)
```

`write_delim` is a generic writer that allows you to set the delimiter.


## reshaping & tidying data

```{r, echo=F}
readr::read_csv("../data/tycho_wide_timeOnly.csv") %>% 
  dplyr::filter(state == "NORTH CAROLINA",
                disease == "MEASLES") %>% 
  readr::write_csv("../data/tycho_nc_measles_wide.csv")

readr::read_csv("../data/tycho_wide_timeOnly.csv") %>% 
  dplyr::filter(state == "TEXAS",
                disease == "MEASLES") %>% 
  readr::write_csv("../data/tycho_tx_measles_wide.csv")
```

```{r, eval=F}
tx_measles_link <- "https://raw.githubusercontent.com/UNC-HLC-R-Intro/lessons_2019/master/data/tycho_tx_measles_wide.csv"
```
```{r, echo=F}
tx_measles_link <- "../data/tycho_tx_measles_wide.csv"
```

```{r}
tx_measles <- read_csv(tx_measles_link)
```
This dataset tracks disease occurrance over time
```{r}
tx_measles 
```

## Exercise 2
How would you make a plot of these data?
```{r, echo=F}
# A: Data must be reformatted for ggplot!
```


```{r}
library(reshape2)
```

Converting data from wide to long (sometimes called "tidy") format is called a "melt" or a "pivot".

The `reshape2` library allows us to melt data
```{r}
melt(tx_measles)
```
## Exercise 3
Read the help documentation for `melt`. How can you set the names of the `variable` and `value` columns?

**Hint:** `melt` has different behavior for different data types, you might have to go to a second help page under "details".

```{r, echo=F}
melt(tx_measles, id.vars = c("state", "disease"), variable.name = "yearweek", value.name = "incidence") %>% head

tx_measles_melt <- melt(tx_measles, id.vars = c("state", "disease"), variable.name = "yearweek", value.name = "incidence")
```

```{r, echo=F}
tx_measles_melt %>% 
  #dplyr::mutate(yearweek = as.numeric(yearweek)) %>% 
  dplyr::mutate(yearweek = as.character(yearweek) %>% as.numeric) %>% 
  ggplot(aes(yearweek, incidence)) +
    geom_line() +
    geom_vline(xintercept = 196301, color = "red")
```


## Full tycho example

```{r}
tycho_link <- "https://raw.githubusercontent.com/UNC-HLC-R-Intro/lessons_2019/master/data/tycho_wide_time-disease.csv"
```

```{r}
tycho <- read_csv(tycho_link)
```

This dataset tracks diseases over time 
```{r}
head(tycho)
```

## Exercise
Tidy this dataset with reshape2

## Tidyr provides functions to manipulate data into tidy format

```{r}
library(tidyr)
```

## Exercise use tidyr to split year and disease