---
title: "Data Import"
output: github_document
always_allow_html: yes
---

```{r setup}
library(tidyverse)
```

Today we will discuss how to import raw data into R.

```{r}
# Use Shift+Click to follow this link to see the raw data
tycho_link <- "https://raw.githubusercontent.com/UNC-HLC-R-Intro/lessons/import/data/import_tycho_simple.csv?token=AQdxTW69CSAGG6JPYwAYDEZ3ze4-AHWRks5bWeAAwA%3D%3D"
tycho_base <- read.csv(tycho_link)
```
```{r, eval=F}
tycho_base %>% head
```
```{r, echo=F}
tycho_base[1:10,1:10]
```


```{r}
tycho_readr <- read_csv(tycho_link)
```
```{r, eval=F}
tycho_readr %>% head
```
```{r, echo=F}
tycho_readr[1:10,1:10]
```


## Exercise 1: base R vs. readr
What are the differences between `tycho_base` and `tycho_readr`?
- What is the structure of the data?
- Are the columns the same type?
- Which function is faster?

## Exercise 2:
- Look at this raw data on github and import it using `readr`.
- What happens?
- What could be a way to fix this problem? (Don't worry about the code solution yet, brainstorm with a partner)
- Can you find a setting in `?read_csv` that might implement your solution(s)?


```{r}
tycho_link_2 <- "https://raw.githubusercontent.com/UNC-HLC-R-Intro/lessons/import/data/import_tycho.csv?token=AQdxTWPiN68L7-gEprtUcHGm257fEeK-ks5bWeERwA%3D%3D"
```

```{r, include=F, echo=F}
tycho2 <- readr::read_csv(tycho_link_2, comment = "#", guess_max = 5)
head(tycho2)
```

## Exercise 3: Regular Expressions

[Use This chapter of "R For Data Science" for reference](http://r4ds.had.co.nz/strings.html#matching-patterns-with-regular-expressions)

Using `str_view_all` to show your results, create regular expressions that answer the following questions:


```{r, eval=F}
# You might need to run this
install.packages("htmlwidgets")
```

```{r}
library(stringr)
```

```{r}
names <- c("Alice", "Bob", "Carol", "Dennis", "Frank", "George")
```

### Match all `names` containing the letter "N"
```{r, echo=F}
str_view_all(names, "n")
```

### Match all names containing the letter "A"
```{r, echo=F}
str_view_all(names, "[Aa]")
```

### Match the first letter of every name
```{r, echo=F}
str_view_all(names, "^[:letter:]")
# Also a valid solution, but will also match if the first character is a number:
# str_view_all(names, "^.")
```

```{r}
myVector <- c(10, "twenty", 30, "fourty", "fifty", 60, "")
```

### Match all `myVector` entries that are numbers (not text)
```{r, echo=F}
str_view_all(myVector, "\\d")
```
### Match all text in `myVector`
```{r, echo=F}
str_view_all(myVector, "[:letter:]")
```

## Exercise 4: replacing with regular expressions
Use the `gsub` command to replace things based on regular expressions
- You can check your regex's with `string_view_all`

### Example: Replace "Bob" for your own name
```{r}
myString <- "My name is Bob"
```
```{r, echo = T}
myString
gsub("Bob", "Spencer", myString)
```
### Replace all spaces for "_" 
```{r, echo = F}
gsub(" ", "_", myString)
```
### Delete all spaces
```{r, echo = F}
gsub(" ", "", myString)
```


# Homework
- Finish any remaining regex examples above
- Read and play with the example found [here](https://github.com/UNC-HLC-R-Intro/lessons/blob/import/data_import/data_import_cleaning_example.md) to see how regex's and `tidyr` can be used to clean up a messy dataset



















