---
title: "Data Manipulation"
output: github_document
---

```{r}
library(gapminder)
library(ggplot2)
library(dplyr)
```

## Replicate this plot

**Hints:**

 - this plot uses the following subset of the Gapminder data
 - this plot uses facets to show the different years
 - a linear trendline
 - log10 x-axis
 - altered transparency of the points
 - renamed legend titles
 - axis text is resized to 12
 - axis title text is resized to 14
 - facet titles are resized to 14
 
```{r}
finalExampleData <- dplyr::filter(gapminder, year %in% c(1952, 2007))
```

```{r gapminder plot, fig.height= 4, fig.width= 9, echo = F}
ggplot(finalExampleData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(size = pop, color = continent), alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE, color = "black", lty = 2) +
  facet_wrap(~year) + 
  scale_x_log10() + 
  ylab("Life Expectancy") + 
  xlab("GDP per capita") + 
  guides(color = guide_legend("Continent"), 
         size = guide_legend("Population")) +
  ggtitle("Life Expectancy Increases independently of GDP per cap") +
  theme(strip.text.x = element_text(size = 14),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))
```

## select
```{r}
pop_by_year <- select(gapminder, country, year, pop)

pop_by_year
```


### Exercise 1
Which country in which year has the highest population?
```{r, echo=F, eval=F, include=F}
# A: China in 2007
arrange(gapminder, desc(pop))
```


## filter
 
| Operator | Meaning |
|:--------:|:-------:|
|    ==    |  Equals |
|    !=    |  Not Equals |
|    %in%    |  within vector |
|    > |  greater than |
|    >= |  greater than or equal|
|    < |  less than |
|    =< |  less than or equal|

```{r}
filter(gapminder, country == "United States")
```


```{r}
filter(gapminder, year > 2000)
```

Multiple filters simultaneously (AND logic).

It is usually a good idea to have line breaks.
```{r}
filter(gapminder, 
       year > 2000, 
       country == "United States")
```

## mutate
gdpPercap is adjusted for population (GDP/Population). We can recompute the
original GDP by multiplying gdpPercap by population.
```{r}
mutate(gapminder, GDP = pop*gdpPercap)
```

## pipes
Pipes send the output of 1 command to the 1st input of the next command.
```{r}
gapminder %>% 
  filter(year > 2000)
```

We can string together complicated operations with pipes
```{r}
gapminder %>% 
  mutate(GDP = gdpPercap * pop) %>% 
  filter(year > 2000) %>% 
  filter(continent == "Asia") %>% 
  select(country, GDP, year)
  
```

## arrange
```{r}
gapminder %>% 
  arrange(year)
```

`desc()` will sort in descending order.
```{r}
arrange(pop_by_year, desc(year))
```

## group_by & summarize

Summarize creates new data frames from functions.
```{r}
gapminder %>% 
  summarize(mean(lifeExp))
```

Columns can be created similar to mutate
```{r}
gapminder %>% 
  summarize(mean_lifeExp = mean(lifeExp))
```

Group_by is **silent**
```{r}
gapminder %>% 
  group_by(country)
```

summarize acts on groups
```{r}
gapminder %>% 
  group_by(country) %>% 
  summarize(mean(lifeExp))
```




```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(meanExp = mean(lifeExp)) 
```

## group_by & mutate

```{r}
gapminder %>% 
  group_by()
```


## Exercise 2
What is the mean life expectancy in each continent for the year 1967?
```{r, echo=F, eval=F}
# Best answer:
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(mean_lifeExp = mean(lifeExp)) %>% 
  filter(year == 1967)

# could also do:   
gapminder %>% 
  filter(year == 1967) %>% 
  group_by(continent, year) %>% 
  summarize(mean_lifeExp = mean(lifeExp))
```

## Exercise 3
Make a new variable `expOver70` that determines whether lifeExpectancy is greater than 70 years old.

Which country/countries was the first to have a life expectancy over 70?
```{r, echo=F, eval=F}
gapminder %>% 
  mutate(expOver70 = lifeExp > 70) %>% # make new column
  filter(expOver70 == T) %>%           # select only areas with lifeExp over 70
  filter(year == min(year)) %>%        # select rows w/ smallest year value (these are the first)
  arrange(desc(lifeExp))

# Alternate solution
gapminder %>% 
  mutate(expOver70 = lifeExp > 70) %>% 
  filter(expOver70 == T) %>%           
  arrange(year, desc(lifeExp))
```


Make a bar plot showing the fraction of countries with life expectancy over 70 over time.

**Bonus:** 

 - Make the bar plot edges touch x axis
 - Remove x axis ticks
 - Resize x axis text
```{r, echo=F}
gapminder %>% 
  mutate(expOver70 = lifeExp > 70) %>% 
  ggplot(aes(factor(year))) +
    geom_bar(aes(fill = expOver70), position = "fill") +
    labs(x = "Year", 
         y = "Fraction of Countries",
         fill = "Life Expectancy > 70") +
    scale_y_continuous(expand = expand_scale(mult = c(0, 0.05))) +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_text(size = 10))

```

## Exercise 4: 
Retrieve country with highest GDP (hint: add a new column and name it GDP)
```{r, echo=F, eval=F}
gapminder %>% 
  mutate(GDP = pop*gdpPercap) %>% 
  filter(GDP == max(GDP))
```
List continents with higest overall GDP from greatest to least
```{r, echo=F, eval=F}
gapminder %>% 
  mutate(GDP = pop*gdpPercap) %>% 
  group_by(continent) %>% 
  summarize(total_gdp = sum(GDP)) %>% 
  arrange(desc(total_gdp))
```
