---
title: "Final Project: Mouse Cortex Protein Expression"
output: github_document
---

## Answer the following questions about the mouse cortex expression data, export your results as a github_document.

```{r setup, warning=F, message=F}
library(tidyverse)
library(magrittr)
```

```{r opts, include=F, warning=F, echo=F}
knitr::opts_chunk$set(echo = F)
```

[Explanation of Data](https://archive.ics.uci.edu/ml/datasets/Mice+Protein+Expression)
```{r}
cortex_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00342/Data_Cortex_Nuclear.xls"

download.file(destfile = "cortex.xls", url = cortex_url)

cortex <- readxl::read_xls("cortex.xls")
```

### Raw Data:
```{r}
cortex[1:4,1:4]
```
## Transform the raw data into a melted data.frame
```{r, include=F}
fct_asInt <- function(x){
  as.character(x) %>% as.numeric(.)
}

cortex_melt <- cortex %>% 
  reshape2::melt(id.vars = c("MouseID", "Treatment", "Behavior", "class", "Genotype"), 
                 variable.name = "protein", 
                 value.name = "expression") %>% 
  dplyr::mutate(expression = fct_asInt(expression))
```

### Melted Data:
```{r, include=T}
head(cortex_melt)
```

## Write a function that will tidy your melted data
 - rename "MouseID" column to "id"
 - rename all column names as lowercase
 - remove the "_N" from the end of all protein names
 - rename the "Control" genotype to "Wild-Type"

```{r}
tidy_data <- function(df){
  
  names(df) <- names(df) %>% 
    tolower() %>% 
    gsub("mouse", "", .)
  
  df <- df %>% 
    dplyr::mutate(protein = gsub("_N", "", protein)) %>% 
    dplyr::mutate(genotype = gsub("Control", "Wild-Type", genotype)) %>% 
    dplyr::mutate(genotype = fct_rev(genotype))
   
  return(df)
}

cortex_melt %<>% 
  tidy_data()
```

### Example of output:
```{r, echo=F, include=T}
head(cortex_melt)
```


```{r}
cortex_mean <- cortex_melt %>% 
  tidyr::drop_na(expression) %>% 
  dplyr::group_by(genotype, treatment, behavior, class, protein) %>% 
  dplyr::summarise(expression = mean(expression)) 
```

```{r, include=F, eval=F}
cortex_mean %>% 
  ggplot(aes(class, log2(expression))) +
    geom_boxplot()
```

## Write a function that plots a ggplot boxplot comparing genotype on the x-axis, log2(expression) on the y-axis where each boxplot is colored by drug/nodrug treatment

- Customize the theme of the plot to your liking. 
- At a minimum, pick non-default colors for your conditions, and resize text labels 
- **Bonus:** Put Wild-Type first in the plot

```{r, fig.height=5, fig.width=5}
trt_boxplot <- function(df){
  ggplot(df, aes(genotype, log2(expression))) +
    geom_boxplot(aes(fill = fct_rev(treatment)), notch = T) +
    scale_fill_manual(values = c("Saline" = "Grey40", "Memantine" = "firebrick")) +
    guides(fill = guide_legend(title = "Treatment")) +
    xlab(NULL) +
    theme(axis.text.x = element_text(size = 14, color = "black"), 
          axis.text.y = element_text(size = 14, color = "black"),
          axis.title.y = element_text(size = 16, color = "black"),
          axis.ticks.x = element_blank(),
          axis.line = element_line(color = "black"), 
          #rect = element_blank(), 
          #panel.grid.major.y = element_line(color = "black"),
          #panel.grid.minor.y = element_line(color = "black"),
          #panel.grid.major.x = element_blank()
          )
    
}

cortex_mean %>% 
  trt_boxplot 
```

## Using group_by, summarise, and tidyr::spread(), rank all proteins by their fold-change in response to drug overall.

### Q: What are the top 5 proteins that increase following drug treatment?
**Hint:** You'll first want to use tidyr::drop_na() to remove missing data or else you'll get `r NA` values!

Your output should look something like this:

(But "protein_#" will be an actual protein name)
```{r}
fc_rank <- cortex_melt %>% 
  tidyr::drop_na() %>% 
  dplyr::group_by(treatment, protein) %>% 
  dplyr::summarise(expression = mean(expression)) %>% 
  tidyr::spread(treatment, expression) %>% 
  dplyr::mutate(fold_change = Memantine / Saline) %>% 
  dplyr::arrange(desc(fold_change))
fc_rank %>% 
  dplyr::mutate(protein = glue::glue("protein_{x}", x = seq_along(protein)))
```

#### Top 5 Increasing proteins:
```{r, include=F}
fc_rank$protein[1:5] 
```

#### Top 5 Decreasing proteins:
```{r, include=F}
rev(fc_rank$protein)[1:5] 
```

## Use the grep command to find any proteins that match histone H3 ("H3")
**Bonus:** Are there any other histones in this dataset? Use regular expressions to find them.
```{r, include=F}
cortex_melt$protein %>% 
  grep("H\\d", ., value = T) %>% 
  unique
# Only H3 is in this dataset, 3 PTM versions are present.
```

## Using your boxplot function, plot the expression level of all H3 proteins (in aggregate) for WT vs Mut, Drug/noDrug
**Hint:** The `grepl` function can be combined with `filter` to select things that match a regex.

### Q: What appears to be the trend?
```{r, include=F}
cortex_melt %>% 
  dplyr::filter(grepl("H\\d", protein)) %>% 
  trt_boxplot() +
    ggtitle("Expression of Histone H3 in response to drug")
```  

## Make a boxplot for each histone protein comparing Wild-Type to Mutant drug/nodrug
**Hint:** Try using facets
**Bonus:** use regex & backreference to rename histone marks in conventional format (ie "H3K9Ac" instead of "AcetylH3K9". Format = "<Histone#><amino-acid#><modification>")

```{r, fig.height=5, fig.width=7.5, include=F}
cortex_melt %>% 
  dplyr::filter(grepl("H\\d", protein)) %>% 
  dplyr::mutate(protein = gsub("Acetyl(.+)", "\\1Ac", protein)) %>% 
  dplyr::mutate(protein = gsub("(H3)(.+)(K\\d+)", "\\1\\3\\2", protein)) %>% 
  trt_boxplot() +
    facet_wrap(~protein) +
    ggtitle("Levels of modified H3 in response to drug") +
    theme(strip.text.x = element_text(size = 14), title = element_text(size = 16))
```

### Q: What is the effect of drug treatment on AcetylH3K9?
(Although you would do stats in real life, don't worry about that now)

```{r, include=F}
# A: 
# Ts65Dn mice have much higher levels of H3K9Ac than Wild-type before drug treatment. 
# While drug treatment has little effect in WT on H3K9Ac levels, drug treatment reduces the level of H3K9Ac.
```