---
title: "Intro to ggplot2"
author: "Spencer Nystrom"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

### Expectations
1. Understand basic `ggplot` syntax
2. Know how `ggplot` views factors and how to relevel them for plotting

Install the `gapminder` dataset
```{r, eval=F, echo=T}
install.packages("gapminder")
```

Next we will load the sample data and subset it. We will cover how to do this in
detail next week, but for now run the following code to generate a `data.frame`
tracking life-expectancy, population, & gdp over time in the United States.
```{r}
library(gapminder)
# We'll cover dplyr next week!
usa <- dplyr::filter(gapminder, country == "United States")
```

Don't forget to load the `ggplot2` library
```{r}
library(ggplot2)
```

## Exercise 1
 - Recall last class's lesson on ggplot. Try to recreate the following scatterplot of life expectancy over time:
 
```{r, echo=F}
ggplot(data = usa, mapping = aes(x = year, y = lifeExp)) +
  geom_point()
```


## Exercise 2
 - Change the point shape to diamonds
 - increase the point size to 5
```{r, echo = F}
# Recall that setting parameters not-informed by data is done **outside** of an aes() call.
ggplot(data = usa, mapping = aes(x = year, y = lifeExp)) +
  geom_point(shape = 18, size = 5)
```

## Layered grammar of graphics

ggplot adds things in layers. The order of addition matters! Lets add a line to
our scatterplot. Observe that the line is plotted over the points.

I've increased the line thickness with the `size` argument so it renders well on a projector.
```{r}
ggplot(data = usa, mapping = aes(x = year, y = lifeExp)) +
  geom_point(shape = 18, size = 5) +
  geom_line(color = "red", size = 1)
```

## Exercise 3
 - redo the above plot so that the points are on top of the line.
```{r, echo=F}
ggplot(data = usa, mapping = aes(x = year, y = lifeExp)) +
  geom_line(color = "red", size = 1) +
  geom_point(shape = 18, size = 5) 
```
 
## Exercise 4
 - set the point size to scale with the population (`pop` column)
 - what happens if you set the color to scale with the population?
 - what happens if you set the color to distinguish the continent?
```{r, echo = F}
ggplot(data = usa, mapping = aes(x = year, y = lifeExp)) +
  geom_point(shape = 18, aes(size = pop))
```

```{r, include=F, echo = F}
# Pt 2&3 solution
ggplot(data = usa, mapping = aes(x = year, y = lifeExp)) +
  geom_point(shape = 18, aes(color = pop))

ggplot(data = usa, mapping = aes(x = year, y = lifeExp)) +
  geom_point(shape = 18, aes(color = continent))
```


# Data Types in ggplot2

Recall the following example:
```{r}
ggplot(iris, aes(Sepal.Width, Petal.Width)) +
  geom_point(aes(color = Species))
```

ggplot interprets categorical variables as `factors` in order to assign their order in the legend, and the color on the plot.


### Factors are ordered categorical data with "Levels"

By default the levels are in **alphabetical order**.

Note how in the below example the 'levels' of `myFactors` are arranged alphabetically.
```{r}
myFactors <- factor(c("One", "Two", "Two", "Three"))

myFactors
```

In the iris dataset, `Species` is already a factor
```{r}
iris$Species
```

## Factors can have their order set on assignment
```{r}
myFactors2 <- factor(c("One", "Two", "Two", "Three"), levels = c("One", "Two", "Three"))

myFactors2
```

## Factors can be reordered after the fact
Use the `forcats` library! It's built for dealing with factors!
```{r}
library(forcats)

myFactors
```

all forcats functions start with `fct_`

`fct_relevel` lets you reorder factors
```{r}
fct_relevel(myFactors, c("One", "Two", "Three"))
```

`fct_rev` reverses the order of levels
```{r}
myFactors2
```
```{r}
fct_rev(myFactors2)
```

## ggplot customization
```{r}
ggplot(iris, aes(Sepal.Width, Petal.Width)) +
  geom_point(aes(color = Species))
```

The order of the factors in the legend (and thus the default color assignment) can be changed by reordering the factor levels!
```{r}
ggplot(iris, aes(Sepal.Width, Petal.Width)) +
  geom_point(aes(color = fct_rev(Species)))
```


## Setting axis labels & title
For these examples I'm going to use the whole gapminder dataset. To save typing I will save a scatterplot of year vs lifeExp to `myPlot`.
```{r}
myPlot <- ggplot(data = gapminder, mapping = aes(x = year, y = lifeExp)) +
    geom_point()

myPlot
```

Titles can be added with `ggtitle()`
```{r}
myPlot +
  ggtitle("Year vs Life Expectancy")
```

Axis titles can be customized with `xlab()` and `ylab()`
```{r}
myPlot +
  ggtitle("Year vs Life Expectancy") +
  xlab("Year") +
  ylab("Life Expectancy")
```

```{r}
myPlot +
  ggtitle("Year vs Life Expectancy") +
  xlab("Year") +
  ylab("Life Expectancy")
```

global aesthetics can be added to the whole plot
```{r}
myPlot +
  aes(color = continent)
```



## Final Exercise

To check your understanding of `ggplot` and topics covered in this lesson,
replicate the plot below of data from the Gapminder dataset. Make use of your
resources in this document, the `ggplot2` documentation, R for Data
Science, and Google.

**Hints:**
 - this plot uses the following subset of the Gapminder data
 - this plot uses facets to show the different years
 - a linear trendline
 - log10 x-axis
 - altered transparency of the points
 - renamed legend titles
 - axis text is resized to 12
 - axis title text is resized to 14
 - facet titles are resized to 14
 

```{r}
finalExampleData <- dplyr::filter(gapminder, year %in% c(1952, 2007))
```

```{r gapminder plot, fig.height= 4, fig.width= 9, echo = F}
ggplot(finalExampleData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(size = pop, color = continent), alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE, color = "black", lty = 2) +
  facet_wrap(~year) + 
  scale_x_log10() + 
  ylab("Life Expectancy") + 
  xlab("GDP per capita") + 
  guides(color = guide_legend("Continent"), 
         size = guide_legend("Population")) +
  ggtitle("Life Expectancy Increases independently of GDP per cap") +
  theme(strip.text.x = element_text(size = 14),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))
```






