---
title: "Class Introduction"
author: "Spencer Nystrom & Grace DiCecco"
output: github_document
---

## Login to Github

### Clone student repo to Rstudio.cloud
 `install.packages("tidyverse")`
 
### Terminal:
  `git config --global user.name <User>`
  
  `git config --global user.email <user@email>`
  
### Create new folder for day1
### Talk about working directories

## Introduce markdown
 - output: github_document
 - knit

## talk about commit and push 
 - save, knit, and commit files to git
 - push changes to github
 - Check github to see output


# Data in R

```{r}
titanic <- as.data.frame(Titanic)
head(titanic)
```

```{r}
library(ggplot2)
```

```{r}
ggplot(data = titanic, aes(x = Class, y = Freq)) # create the plotting space
```

```{r}
ggplot(data = titanic, aes(x = Class, y = Freq)) +
    geom_col()
```


```{r}
ggplot(titanic, aes(Class, Freq)) +
    geom_col(aes(fill = Survived))
```

```{r}
ggplot(titanic, aes(Class, Freq)) +
    geom_col(aes(fill = Survived), position = "fill")
```

