---
title: "Class Introduction"
author: "Spencer Nystrom & Jamshaid Shahir"
output: github_document
---

## Class Intro
 - Student introductions
 - class expectations:
   - work together
   - ask questions
   - be respectful of others
   - making mistakes is the best way to learn!

## Rstudio.cloud setup

 - login w/ github account
 - disable "restore environemnt" setting in Rstudio
 - explain each window
 

```{r, eval=F}
# This installs multiple packages that help manipulate & plot data:
install.packages("tidyverse")

# This installs two packages that contain example datasets we will use later in the course:
install.packages(c("gapminder", "titanic"))
```

### Create new folder for day1
### Talk about working directories

### R scripts vs. console


# Intro to R

## By default R spits out things to terminal
```{r}
2 + 2

100*100

"My name is Spencer"
```

## You can store things in **variables** with `<-`

You can quickly type `<-` with `Alt + -` in Rstudio

```{r}
myName <- "Spencer"
```

## R will spit it back out when you call it
```{r}
myName
```

## R can do operations on variables
```{r}
myNumber <- 10

myNumber + 10
```

## Comments are hidden from R
```{r}
# Text after a # will not be run, it is a "comment". 
# This lets you document your scripts. Always comment your code!
```



# Data in R

`iris` is a built-in dataset that we will use today to cover some basic. It
contains measurements of 3 different species of flower.

We can look at the contents by printing it to the console (annoying for large datasets)
```{r}
iris
```

The `head()` function will print the first 6 rows
```{r}
head(iris)
```

The `View()` function will open the dataset in Rstudio's data viewer.
```{r, eval=F}
View(iris)
```

Datasets can be stored in R as **data.frames**. `iris` is a data.frame. You can think of these like excel tables.

What is special about data.frames is that each column contains data of a certain type. R sees numbers and words as different things. Number columns can be used in math operations, but word columns cannot.
**Note:** We will talk in more detail about these data types later

You can view the type of each column with the `str()` (Structure) function. Here you can see that `Sepal.Length` is a number (`num`), but `Species` is a type called a `Factor`.
```{r}
str(iris)
```

Individual columns can be referenced in R by using `$` to call a column by name.
```{r}
iris$Sepal.Length
```

```{r}
iris$Species
```

R will do operations along entire columns. To multiply all `Sepal.Length` values by 2, for example:
```{r}
iris$Sepal.Length * 2
```

## Exercise (3 minutes):
 - print all values of `Petal.Width` to the terminal
 - You can ask for help in R using the `?` function. Try running `?mean` to find out what the `mean()` function does.
 - what is the mean value of `Petal.Width`?
 - what is the sum of all `Petal.Width`? (HINT: guess the function name, or google for it!)
 
 
## Plotting

The package `ggplot2` is used for making plots from data.frames. Load the package by calling `library(ggplot2)`. You only have to load libraries once per script.

```{r}
library(ggplot2)
```


The function `ggplot()` is used to build plots from data. By calling `ggplot` alone we build a totally empty plot.

```{r}
ggplot()
```

We tell ggplot to use a dataset by giving it a `data.frame` to the `data` argument.
```{r}
ggplot(data = iris)
```

Lets make a scatterplot of Sepal width vs Petal width.

To do this we have to **map** the data onto the plot. To do this we make an **aesthetic** (with `aes()`). **aesthetics** are parts of the plot that are **determined by the data**. Here we say that we want to plot the `Sepal.Width` on the x-axis, and `Petal.Width` on the y-axis.

This creates a plotting area that is informed by the data in `iris`.
```{r}
ggplot(data = iris, mapping = aes(x = Sepal.Width, y = Petal.Width)) 
```

Now let's add the points. We do this by adding a `geom`. In this case `geom_point` for datapoints.
```{r}
ggplot(data = iris, mapping = aes(x = Sepal.Width, y = Petal.Width)) +
  geom_point()
```

Plots can be saved to variables just like before
```{r}
sepal.vs.petal <- ggplot(data = iris, mapping = aes(x = Sepal.Width, y = Petal.Width)) +
  geom_point()
```

```{r}
sepal.vs.petal
```

We can edit the parameters of geoms. For example, you can change the color of the points by editing the `color` parameter.

```{r}
ggplot(data = iris, mapping = aes(x = Sepal.Width, y = Petal.Width)) +
  geom_point(color = "red")
```

## Exercise:
### Question 1:
 - Try making the same scatterplot but color the points by their `Species`. 
 - **Hints:**
  - Try looking at the help page for `aes()` with `?aes`
  - If this isn't helping, try googling a solution!
  
```{r, eval=F, include =F}
# Correct answer:
ggplot(data = iris, mapping = aes(x = Sepal.Width, y = Petal.Width)) +
  geom_point(aes(color = Species))
```

### Question 2:
 - Make a boxplot like the one below plotting the `Petal.Width` for each `Species`
 - **Hints:**
  - which `geom` adds a boxplot?
  - How do you fill-in the boxplot by species?
  - Try googling a solution.

```{r, echo=F}
# Correct answer:
ggplot(data = iris, mapping = aes(x = Species, y = Petal.Width)) +
  geom_boxplot(aes(fill = Species)) 
```

## More reading
[R for Data Science: ggplot2 introduction](https://r4ds.had.co.nz/data-visualisation.html)
 - Great textbook!

[ggplot2 reference manual](https://ggplot2.tidyverse.org/reference/)
 - really useful manual with pictures!